% !TEX TS-program = lualatex

\DocumentMetadata{
  lang = en,
  pdfversion = 2.0,
  pdfstandard = ua-2,
  tagging = on,
  tagging-setup = {
    math/setup = {mathml-SE},
    table/header-rows = 1
  }
}

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{geometry}
\geometry{
  letterpaper,
  textwidth=6.5in,
  textheight=9in,
  top=0.85in,
  bottom=1in,
  left=0.75in,
  right=0.75in,
}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\textbf{#1.} }{\ \rule{0.5em}{0.5em}}

\tagpdfsetup{
  math/alt/use,
  role/new-tag=section/H1,
}

\hypersetup{
  pdftitle={Assignment 6 -- Key},
  pdfauthor={Blankenau},
  pdfsubject={Fall, 2025},
  hidelinks,
}

\begin{document}
\begin{center}
{\LARGE Assignment 6}
\end{center}

\noindent Blankenau

\noindent Economics 905, Fall, 2025

\medskip
\noindent \noindent \textit{Instructions.} You may work in groups and may
compare answers prior to submission. Joint submissions with up to 4 names
are allowed and encouraged. All work should be legible and concise. A due
date will be announced in class.

\begin{enumerate}
\item Describe how each of the following affects the $\dot{k}=0$ and $\dot{c}%
=0$ curves in our phase diagram, and thus how they affect the
balanced-growth-path values of $c$ and $k$:

\begin{enumerate}
\item A rise in $\theta .$

\item A change in the depreciation rate from the value of 0 assumed so far
to some positive level. That is, capital now evolves according to $\overset{.%
}{k}\left( t\right) =f\left( k\left( t\right) \right) -c\left( t\right)
-\left( n+g+\delta \right) k\left( t\right) .$ Note in addition that now the
return on capital is $r\left( t\right) =f^{\prime }\left( k\left( t\right)
\right) -\delta \ $rather than $r\left( t\right) =f^{\prime }\left( k\left(
t\right) \right) .$ This should lead you to conclude that both the $\dot{k}%
=0 $ and $\dot{c}=0$ shift.
\end{enumerate}

\item Consider a standard Ramsey-Cass-Koopman's growth model where
investment income is subsidized at rate $\varsigma _{k}.$ In each period,
revenue for the subsidy is generated from a lump sum tax of $G\left(
t\right) $ per effective unit of labor. The representative household
maximizes 
\begin{equation*}
U=\dint\limits_{t=0}^{\infty }e^{-\left( \rho -n-\left( 1-\theta \right)
g\right) t}\frac{c\left( t\right) ^{1-\theta }}{1-\theta }dt
\end{equation*}%
$\theta >0,$ $\ \rho -n-\left( 1-\theta \right) g>0$ subject to%
\begin{equation*}
k\left( 0\right) +\dint\limits_{t=0}^{\infty }e^{-R\left( t\right)
}e^{\left( n+g\right) t}\left( w\left( t\right) -G\left( t\right) \right)
dt=\dint\limits_{t=0}^{\infty }e^{-R\left( t\right) }e^{\left( n+g\right)
t}c\left( t\right) dt
\end{equation*}%
and the standard transversality condition. The capital stock per effective
unit of labor evolves according to $\overset{.}{k}\left( t\right) =f\left(
k\left( t\right) \right) -c\left( t\right) -\left( n+g\right) k\left(
t\right) .$ All items are defined as in the text: $c\left( t\right) $ and $%
k\left( t\right) $ are consumption and capital per effective unit of labor, $%
n$ in the growth rate of the population, $g$ is the growth rate of the labor
augmenting technology, $\rho $ is the discount rate, $R\left( t\right)
=\dint\limits_{\tau =0}^{t}r\left( \tau \right) d\tau $ $\ $where $r\left(
\tau \right) $ is the period $\tau $ interest rate, $w\left( t\right) $ is
the period $t$ wage. Thus the real interest rate facing households is given
by $\left( 1+\varsigma _{k}\right) f^{\prime }\left( k\left( t\right)
\right) .$ The intensive form production function $f\left( k\left( t\right)
\right) $ has the standard properties.

\begin{enumerate}
\item Show your derivation of the optimal growth rate of consumption.

\item Draw the $\overset{.}{k}\left( t\right) =0$ and $\overset{.}{c}\left(
t\right) =0$ loci in the $c,\,k$ space (that is, draw the principal curves
of the phase diagram). Label everything properly. Label the steady state
level of capital $k^{\ast }.$

\item For this part only let $n=g=0$ and $f\left( k\left( t\right) \right)
=y\left( t\right) =\left( k\left( t\right) \right) ^{.5}$. Suppose that the
economy has been operating at $k\left( t\right) =\frac{K_{0}}{L_{0}A_{0}}%
=k_{0}.$ Suddenly $k\left( t\right) $ changes to $k=zk_{0}\ $where $z<1$
This might have happened because of destruction of some of the capital
stock, a one-time permanent increase in $L$ or a one-time permanent increase
in $A$. A policy maker says that it does not matter what caused this
decrease since all dynamics are the same, regardless of the cause of the
decrease. Do you agree with this? If so, explain why. If not, describe the
differences across the three cases, paying particular attention to $y\left(
t\right) $ and per capita output. Consider these items both immediately and
in the long run.

\item Suppose the economy is again at $k=k^{\ast }.$ What are the immediate
and long run consequence on $c$ and $k$ of an increase in $\varsigma _{k}?$
Show this on a phase diagram.

\item A policy maker says that in this situation, the effect of an increase
in $\varsigma _{k}=0$ to $\varsigma _{k}>0$ on utility, $U$, is ambiguous
because consumption is affected differently in the short run and in the long
run. Do you agree with this statement? Why or why not?\end{enumerate}
\end{enumerate}

\section*{Answer Key}

\begin{enumerate}
\item \textit{See below.}
\begin{enumerate}
\item \textbf{For all of this, be sure to draw the pictures. I only
describe them here and can answer questions in class.} From 
\begin{eqnarray*}
\dot{k}\left( t\right)  &=&f\left( k\left( t\right) \right) -c\left(
t\right) -\left( n+g\right) k\left( t\right)  \\
\frac{\dot{c}\left( t\right) }{c\left( t\right) } &=&\frac{f^{\prime }\left(
k\left( t\right) \right) -\rho -\theta g}{\theta }
\end{eqnarray*}%
we know that the curves are  
\begin{eqnarray*}
\dot{k}\left( t\right)  &:&c\left( t\right) =f\left( k\left( t\right)
\right) -\left( n+g\right) k\left( t\right)  \\
\dot{c}\left( t\right)  &:&f^{\prime }\left( k\left( t\right) \right) =\rho
+\theta g
\end{eqnarray*}%
The parameter $\theta $ appears only in $f^{\prime }\left( k\left( t\right)
\right) =\rho +\theta g$ so only this curve moves. With $\rho +\theta g$
larger, $f^{\prime }\left( k\left( t\right) \right) $ must be larger which
occurs at a smaller $k$. Thus this curve shifts left. The new intersection
is at a smaller $c$ and $k$ so steady state values of these are lower.
However, the immediate effect of $c\left( t\right) $ is a jump up to get on
the new saddle path. $k(t)$ does not change immediately. Note that your new
saddle path should be above the $\dot{k}\left( t\right) =0$ curve. 

\item Now 
\begin{eqnarray*}
\dot{k}\left( t\right)  &:&c\left( t\right) =f\left( k\left( t\right)
\right) -\left( n+g+\delta \right) k\left( t\right)  \\
\dot{c}\left( t\right)  &:&f^{\prime }\left( k\left( t\right) \right) =\rho
+\theta g+\delta 
\end{eqnarray*}
Since $\delta $ is in each, both curves move. We consider them one at a
time. 

\begin{itemize}
\item Your $\dot{k}\left( t\right) =0$ curve should shifts down. In
particular, it intersects the horizontal axis at the same point on the left,
but at a smaller amount of capital to the right so the new curve is
everywhere below the original curve. From this effect in isolation, the
curves intersect at the same capital and less consumption. There is an
immediate jump to the new level of consumption. 

\begin{itemize}
\item Immediate effects are less consumption, same capital. 

\item Long run effect are less consumption, same capital. 
\end{itemize}

\item Now consider the other shift in isolation. With $\rho +\theta g+\delta 
$ larger, $f^{\prime }\left( k\right) $ must be larger. This occurs with
less $k.$ The curve shifts to the left. 

\begin{itemize}
\item Immediately, consumption increases to get on the new saddle path. 

\item In the long run, capital falls and consumption falls since the new
intersection is at lower $c$ and $k$. 
\end{itemize}

\item So immediately, we have downward pressure on consumption from the
first shift and upward from the second shift. We do not know the effect on
immediate consumption because we do not know which of these dominates.
Immediately, neither shift moves capital so capital does not change. 

\item In the long run we have downward pressure on consumption from both
effects so it definitely falls. In the long run we have no change in capital
from the first effect and a downward change from the other. So $k$
definitely falls in the long run. 
\end{itemize}
\end{enumerate}

\item \begin{enumerate}
\item except for $r_{t}=\left( 1+\varsigma _{k}\right) f^{\prime }\left(
k\left( t\right) \right) ,$ this is the same as in your notes. 

\item as in notes. 

\item We know how to handle the case of larger $A$ from your notes, just
generalizing from $1.1$ to $z^{-1}$, the relevant rows are     
\begin{table}[htbp]
\centering
\begin{tabular}{llcc}
Item & Prior to shock & Immediate & Long Run \\ 
$y$ & \multicolumn{1}{|l}{$y_{0}=k_{0}^{.5}=\left( \frac{K_{0}}{A_{0}L}%
\right) ^{.5}$} & \multicolumn{1}{|c}{$\left( \frac{K_{0}}{A_{1}L}\right)
^{.5}=\left( \frac{1}{z^{-1}}\right) ^{.5}k_{0}^{.5}=z^{0.5}y_{0}$ } & 
\multicolumn{1}{|c}{$y_{1}=y_{0}$} \\ 
$\frac{Y}{L}$ & \multicolumn{1}{|l}{$A_{0}y_{0}$} & \multicolumn{1}{|c}{$%
z^{0.5}A_{1}y_{0}=z^{-0.5}A_{0}y_{0}$} & \multicolumn{1}{|c}{$%
z^{-1}A_{0}y_{0}$}%
\end{tabular}
\caption{Table 1: Comparison of values.}
\end{table}
If it is instead an increase in $L.$%
\begin{table}[htbp]
\centering
\begin{tabular}{llcc}
Item & Prior to shock & Immediate & Long Run \\ 
$y$ & \multicolumn{1}{|l}{$y_{0}=k_{0}^{.5}=\left( \frac{K_{0}}{A_{0}L}%
\right) ^{.5}$} & \multicolumn{1}{|c}{$\left( \frac{K_{0}}{AL_{1}}\right)
^{.5}=\left( \frac{1}{z^{-1}}\right) ^{.5}k_{0}^{.5}=z^{0.5}y_{0}$ } & 
\multicolumn{1}{|c}{$y_{1}=y_{0}$} \\ 
$\frac{Y}{L}$ & \multicolumn{1}{|l}{$Ay_{0}$} & \multicolumn{1}{|c}{$%
z^{0.5}Ay_{0}$} & \multicolumn{1}{|c}{$Ay_{0}$}%
\end{tabular}
\caption{Table 2: Comparison of values.}
\end{table}
If it is instead $K$ 
\begin{table}[htbp]
\centering
\begin{tabular}{llcc}
Item & Prior to shock & Immediate & Long Run \\ 
$y$ & \multicolumn{1}{|l}{$y_{0}=k_{0}^{.5}=\left( \frac{K_{0}}{A_{0}L}%
\right) ^{.5}$} & \multicolumn{1}{|c}{$\left( \frac{zK_{0}}{AL}\right)
^{.5}=z^{0.5}y_{0}$ } & \multicolumn{1}{|c}{$y_{1}=y_{0}$} \\ 
$\frac{Y}{L}$ & \multicolumn{1}{|l}{$Ay_{0}$} & \multicolumn{1}{|c}{$%
z^{0.5}Ay_{0}$} & \multicolumn{1}{|c}{$Ay_{0}$}%
\end{tabular}
\caption{Table 3: Comparison of values.}
\end{table}A graphical and intuitive response will also suffice. 

\item In this case there is no government consumption so the $\dot{k}\left(
t\right) =0$ curve does not move. But now%
\begin{equation*}
\dot{c}\left( t\right) :f^{\prime }\left( k\left( t\right) \right) \left(
1+\varsigma _{k}\right) =\rho +\theta g+\delta 
\end{equation*}%
An increase in $\varsigma _{k}$ increases the left hand side but does not
change the right. $f^{\prime }\left( k\left( t\right) \right) $ must
decrease to preserve the equality. This happens at higher $k.$ The curve
shifts right. Immediately consumption falls to get on the new saddle path.
In the long run $c$ and $k$ increase.   

\item In class.  
\end{enumerate}
\end{enumerate}
\end{document}
