%2multibyte Version: 5.50.0.2960 CodePage: 1252

\documentclass[11pt]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsmath}

\setcounter{MaxMatrixCols}{10}
%TCIDATA{OutputFilter=LATEX.DLL}
%TCIDATA{Version=5.50.0.2960}
%TCIDATA{Codepage=1252}
%TCIDATA{<META NAME="SaveForMode" CONTENT="1">}
%TCIDATA{BibliographyScheme=Manual}
%TCIDATA{Created=Monday, September 22, 2003 09:40:11}
%TCIDATA{LastRevised=Tuesday, November 11, 2025 12:10:47}
%TCIDATA{<META NAME="GraphicsSave" CONTENT="32">}
%TCIDATA{<META NAME="DocumentShell" CONTENT="General\SW\Blank - Standard LaTeX Article">}
%TCIDATA{Language=American English}
%TCIDATA{CSTFile=LaTeX article (bright).cst}

\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\textbf{#1.} }{\ \rule{0.5em}{0.5em}}
\input{tcilatex}
\input{tcilatex}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9in}
\setlength{\topmargin}{-.50in}
\setlength{\oddsidemargin}{.0in}
\setlength{\evensidemargin}{.0in}

\begin{document}


\begin{center}
{\LARGE Assignment 9}
\end{center}

\noindent Blankenau\hspace*{\stretch{0.7500}}

\noindent Economics 905, Fall 2025

\noindent \noindent \textit{Instructions.} You may work in groups and may
compare answers prior to submission. Joint submissions with up to 4 names
are allowed and encouraged. All work should be legible and concise. A due
date will be announced in class.

\begin{enumerate}
\item Consider the following economy. The representative agent maximizes%
\begin{equation*}
E_{0}\overset{\infty }{\underset{t=0}{\sum }}\beta ^{t}\left( \ln
c_{t}\right)
\end{equation*}%
where $0<\beta <1,$ and $c_{t}>0$ is consumption. The representative agent
has one unit of time available in each period. Output is produced according
to 
\begin{equation*}
y_{t}=z_{t}\bar{k}_{t}^{\gamma }k_{t}^{\alpha }n_{t}^{1-\alpha }
\end{equation*}%
where $k_{t}$ is the capital input, $n_{t}$ is the labor input, and $%
0<\alpha <1.$ This is a learning-by-doing economy so $\bar{k}_{t}$ is the
average capital stock, taken as given by all agents and firms. Assume $%
\gamma +\alpha <1$ so this is not an endogenous growth model. Capital
depreciates fully in each period. Assume that $\ln z_{t}$ is an i.i.d.
random variable with an expected value of $\mu >0$ and that $k_{0}$ is
given. Since leisure is not valued, $n_{t}$ will be one in equilibrium. Note
that relative to your notes, $\rho $ here is 0.

\begin{enumerate}
\item Formulate the stochastic dynamic programming problem for the social
planner. Recall the social planner internalizes this externality and so
works with $y_{t}=z_{t}k_{t}^{\alpha +\gamma }n_{t}^{1-\alpha }.$ This
should not fundamentally change how you solve the problem. It will be easier
to define $\delta \equiv \alpha +\gamma $ and work with $z_{t}k_{t}^{\delta
}n_{t}^{1-\alpha }$ in the social planner's problem. Notice that leisure is
not valued so there is one less first order condition relative to your notes
and $n_{t}=1.$ Guess that the value function takes the form%
\begin{equation*}
\upsilon \left( k_{t},z_{t}\right) =B+D\ln k_{t}+F\ln z_{t}
\end{equation*}%
and verify that this is correct by solving for $D$ and $F$. You do not need
to solve for $B$.

\item Now consider the competitive equilibrium. Formulate the stochastic
dynamic programming problem for the representative agent with budget
constraint $c_{t}+k_{t+1}=w_{t}n_{t}+r_{t}k_{t}.$ Here $w_{t}$ is the wage
and $r_{t}$ is the real interest rate. Guess that the value function takes
the form%
\begin{equation*}
\upsilon \left( k_{t},z_{t},\bar{k}_{t}\right) =B+D\ln k_{t}+F\ln z_{t}+G\ln 
\bar{k}_{t}
\end{equation*}%
and verify that this is correct by finding $D$ and $F,$ and $G$ where $D$
and $F$ are potentially different from what you found above. You do not need
to solve for $B$. Note that after taking first order conditions in this case
you can plug in for $w_{t}n_{t}+r_{t}k_{t}=z_{t}\bar{k}_{t}^{\gamma
}k_{t}^{\alpha }n_{t}^{1-\alpha }\ $(It is straightforward to show this, but
you can take it as given). Notice that since leisure is not valued $n_{t}=1$
in equilibrium$.$ Do not separately set $\bar{k}_{t}=k_{t}$ in this
situation.
\end{enumerate}
\end{enumerate}

\pagebreak \pagebreak

\begin{enumerate}
\item 
\begin{enumerate}
\item The dynamic programming problem for this model is%
\begin{equation*}
v(k_{t},z_{t})=\underset{n_{t},k_{t+1}}{\max }\{\ln \left(
z_{t}k_{t}^{\delta }n_{t}^{1-\alpha }-k_{t+1}\right) +\beta
E_{t}v(k_{t+1},z_{t+1})\}
\end{equation*}%
We guess the value function to be of the form: 
\begin{equation*}
v(k_{t},z_{t})=B+D\ln k_{t}+F\ln z_{t}
\end{equation*}%
Now we plug our guess into the Bellman equation and solve the optimization
problem for $k_{t+1}$ 
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}=\underset{k_{t+1}}{\max }\{\ln \left(
z_{t}k_{t}^{\delta }n_{t}^{1-\alpha }-k_{t+1}\right) +\beta E_{t}\left[
B+D\ln k_{t+1}+F\ln z_{t+1}\right] \}
\end{equation*}%
Since $k_{t+1}$ is something controlled by the planner, it is not
stochastic. Also $B$ is clearly not stochastic and $E_{t}\ln z_{t+1}=\mu $
so this is we 
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}=\underset{k_{t+1}}{\max }\{\ln \left(
z_{t}k_{t}^{\delta }n_{t}^{1-\alpha }-k_{t+1}\right) +\beta \left[ B+D\ln
k_{t+1}+F\ln \mu \right] \}
\end{equation*}%
Then, solving the maximization problem on the right-hand side of the Bellman
equation gives the following:%
\begin{equation*}
\frac{1}{z_{t}k_{t}^{\delta }n_{t}^{1-\alpha }-k_{t+1}}=\frac{\beta D}{%
k_{t+1}}
\end{equation*}%
so%
\begin{equation*}
k_{t+1}=\frac{\beta Dz_{t}k_{t}^{\delta }n_{t}^{1-\alpha }}{1+\beta D}
\end{equation*}%
This into above expression gives 
\begin{eqnarray*}
B+D\ln k_{t}+F\ln z_{t} &=&\{\ln \left( z_{t}k_{t}^{\delta }n^{1-\alpha }-%
\frac{\beta D}{\left( 1+\beta D\right) }z_{t}k_{t}^{\delta }n^{1-\alpha
}\right) + \\
&&\beta B+\beta D\ln \frac{\beta D}{\left( 1+\beta D\right) }%
z_{t}k_{t}^{\delta }n^{1-\alpha }+\beta F\mu \}
\end{eqnarray*}%
Now let $n=1$ giving 
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}=\{\ln \left( z_{t}k_{t}^{\delta }\left( 1-\frac{%
\beta D}{\left( 1+\beta D\right) }1\right) \right) +\beta B+\beta D\ln \frac{%
\beta D}{\left( 1+\beta D\right) }z_{t}k_{t}^{\delta }n^{1-\alpha }+\beta
F\mu \}
\end{equation*}%
or%
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}=\{\ln \left( z_{t}k_{t}^{\delta }\right) +\beta D\ln
z_{t}k_{t}^{\delta }+X\}
\end{equation*}%
or 
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}=\{\ln z_{t}+\delta \ln k_{t}+\beta D\delta \ln
k_{t}+\beta D\ln z_{t}+X\}
\end{equation*}%
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}=\{\left( 1+\beta D\right) \ln z_{t}+\left( \delta
+\beta D\delta \right) \ln k+X\}
\end{equation*}%
Equating coefficients: 
\begin{equation*}
D=\delta +\beta D\delta
\end{equation*}%
or 
\begin{equation*}
D=\frac{\delta }{1-\beta \delta }
\end{equation*}%
Similarly 
\begin{equation*}
F=1+\beta D=\frac{1}{1-\beta \delta }
\end{equation*}%
The rest of this was not required but we can also find the policy function 
\begin{equation*}
k_{t+1}=\frac{\beta Dz_{t}k_{t}^{\delta }n_{t}^{1-\alpha }}{1+\beta D}
\end{equation*}%
\begin{equation*}
k_{t+1}=\frac{\beta \frac{\delta }{1-\beta \delta }z_{t}k_{t}^{\delta
}n_{t}^{1-\alpha }}{1+\beta \frac{\delta }{1-\beta \delta }}
\end{equation*}%
becomes 
\begin{equation*}
k_{t+1}=\delta \beta z_{t}k_{t}^{\delta }n_{t}^{1-\alpha }
\end{equation*}

\item The dynamic programming problem for this model is: 
\begin{equation*}
v(k_{t},z_{t},)=\underset{c_{t},n_{t},k_{t+1}}{\max }\{\ln \left(
n_{t}w_{t}+r_{t}k_{t}-k_{t+1}\right) +\beta E_{t}v(k_{t+1},z_{t+1})\}
\end{equation*}%
We guess the value function to be of the form: 
\begin{equation*}
v(k_{t},z_{t})=B+D\ln k_{t}+F\ln z_{t}+G\ln \bar{k}_{t}
\end{equation*}%
Now we will plug our guess into the Bellman equation and solve the
optimization problem for $k_{t+1}$, $n_{t}$: 
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}+G\ln \bar{k}_{t}=\underset{k_{t+1}}{\max }\{\ln
\left( w_{t}+r_{t}k_{t}-k_{t+1}\right) +\beta E_{t}\left[ B+D\ln
k_{t+1}+F\ln z_{t+1}+G\ln \bar{k}_{t+1}\right] \}
\end{equation*}%
Since $k_{t+1}$ is something controlled by the agent, it is not stochastic. $%
\bar{k}_{t+1}$ is not controlled by this agent but is known. Also $B$ is not
stochastic and $E_{t}\ln z_{t+1}=\mu $ so we can write 
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}+G\ln \bar{k}_{t}=\underset{k_{t+1}}{\max }\{\ln
\left( w_{t}n_{t}+r_{t}k_{t}-k_{t+1}\right) +\beta \left[ B+D\ln
k_{t+1}+F\ln \mu +G\ln \bar{k}_{t+1}\right] \}
\end{equation*}%
Differentiation wrt $k_{t+1}$ give 
\begin{equation*}
\frac{1}{w_{t}n_{t}+r_{t}k_{t}-k_{t+1}}=\frac{\beta D}{k_{t+1}}
\end{equation*}%
It was given to you that $w_{t}n_{t}+r_{t}k_{t}=z_{t}\bar{k}_{t}^{\gamma
}k_{t}^{\alpha }n_{t}^{1-\alpha }.$ To see it, we know from firm's problem
that%
\begin{equation*}
r_{t}=\alpha z_{t}\bar{k}_{t}^{\gamma }k_{t}^{\alpha -1}n_{t}^{1-\alpha }
\end{equation*}%
, 
\begin{equation*}
w_{t}=\left( 1-\alpha \right) z_{t}\bar{k}_{t}^{\gamma }k_{t}^{\alpha
}n_{t}^{-\alpha }
\end{equation*}%
so 
\begin{equation*}
w_{t}n_{t}+r_{t}k_{t}=\left( 1-\alpha \right) z_{t}\bar{k}_{t}^{\gamma
}k_{t}^{\alpha }n_{t}^{1-\alpha }+\alpha z_{t}\bar{k}_{t}^{\gamma
}k_{t}^{\alpha }n_{t}^{1-\alpha }=z_{t}\bar{k}_{t}^{\gamma }k_{t}^{\alpha
}n_{t}^{1-\alpha }
\end{equation*}%
Given this, we have 
\begin{equation*}
\frac{1}{z_{t}\bar{k}_{t}^{\gamma }k_{t}^{\alpha }n_{t}^{1-\alpha }-k_{t+1}}=%
\frac{\beta D}{k_{t+1}}
\end{equation*}%
$\allowbreak \allowbreak $%
\begin{equation*}
k_{t+1}=\frac{\beta Dz_{t}\bar{k}_{t}^{\gamma }k_{t}^{\alpha
}n_{t}^{1-\alpha }}{1+\beta D}
\end{equation*}%
\begin{eqnarray*}
B+D\ln k_{t}+F\ln z_{t}+G\ln \bar{k}_{t} &=&\{\ln \left( z_{t}\bar{k}%
_{t}^{\gamma }k_{t}^{\alpha }n_{t}^{1-\alpha }-\frac{\beta D}{\left( 1+\beta
D\right) }z_{t}\bar{k}_{t}^{\gamma }k_{t}^{\alpha }n_{t}^{1-\alpha }\right) +
\\
&&\beta B+\beta D\ln \frac{\beta D}{\left( 1+\beta D\right) }z_{t}\bar{k}%
_{t}^{\gamma }k_{t}^{\alpha }n_{t}^{1-\alpha }+\beta F\mu +\beta G\ln \bar{k}%
_{t+1}
\end{eqnarray*}%
set $n=1$ which we can write as%
\begin{eqnarray*}
B+D\ln k_{t}+F\ln z_{t}+G\ln \bar{k}_{t} &=&\{\ln \left( z_{t}\bar{k}%
_{t}^{\gamma }k_{t}^{\alpha }\left( 1-\frac{\beta D}{\left( 1+\beta D\right) 
}\right) \right) + \\
&&\beta B+\beta D\ln \frac{\beta D}{\left( 1+\beta D\right) }z_{t}\bar{k}%
_{t}^{\gamma }k_{t}^{\alpha }+\beta F\mu +\beta G\ln \bar{k}_{t+1}
\end{eqnarray*}%
or%
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}+G\ln \bar{k}_{t}=\{\ln \left( z_{t}\bar{k}%
_{t}^{\gamma }k_{t}^{\alpha }\right) +\beta D\ln z_{t}\bar{k}_{t}^{\gamma
}k_{t}^{\alpha }+\beta G\ln \bar{k}_{t+1}+X\}
\end{equation*}%
where $X$ is a constant or 
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}+G\ln \bar{k}_{t}=\{\ln z_{t}+\gamma \ln \bar{k}%
_{t}+\alpha \ln k_{t}+\beta D\alpha \ln k_{t}+\beta D\ln z_{t}+\beta D\gamma
\ln \bar{k}_{t}+\beta G\ln \bar{k}_{t}+X\}
\end{equation*}%
\begin{equation*}
B+D\ln k_{t}+F\ln z_{t}+G\ln \bar{k}_{t}=\left( 1+\beta D\right) \ln
z_{t}+\left( \gamma +\beta D\gamma +\beta G\right) \ln \bar{k}_{t}+\left(
\alpha +\beta D\alpha \right) \ln k_{t}+X
\end{equation*}%
where $X$ is a constant. Equating coefficients:%
\begin{equation*}
D=\alpha +\beta D\alpha 
\end{equation*}%
or 
\begin{equation*}
D=\frac{\alpha }{1-\beta \alpha }
\end{equation*}%
Similarly%
\begin{equation*}
F=1+\beta D=\frac{1}{1-\beta \alpha }
\end{equation*}%
$.$%
\begin{equation*}
G=\gamma +\beta D\gamma +\beta G
\end{equation*}%
\begin{equation*}
G=\frac{\gamma \left( 1+D\beta \right) }{1-\beta }
\end{equation*}%
\begin{equation*}
G=\gamma +\beta \frac{\alpha }{1-\beta \alpha }\gamma +\beta G
\end{equation*}%
$\allowbreak $%
\begin{equation*}
G=\frac{\gamma \left( 1+\frac{\alpha }{1-\beta \alpha }\beta \right) }{%
\left( 1-\beta \right) }
\end{equation*}%
\begin{equation*}
G=\frac{\gamma }{\left( 1-\beta \right) \left( 1-\alpha \beta \right) }
\end{equation*}
\end{enumerate}
\end{enumerate}

\end{document}
